---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## 1. Loading and preprocessing the data
1.1. read the CSV file into a dataframe called Activity  
assumption: activity.csv is in the working directory
````{r ,Loading, echo=TRUE}
Activity<-read.csv("activity.csv")
````

## 2. What is mean total number of steps taken per day?
load Dplyr, as this is used to group and summarise the dataframe
````{r , Load Dplyr,echo=TRUE}
library(dplyr)
````

2.1 use Dplyr to group the dataframe by the date coloum and store in variable tmp
````{r ,groupByDate, echo=TRUE}
tmp<-dplyr::group_by(Activity, date) 
````
take the mean of each the steps per day and store in variable tmp2
````{r ,echo=TRUE}
tmp2<-dplyr::summarise_each(tmp, funs(mean))
````
drop the third coloumn as not relevant 
````{r ,echo=TRUE}
tmp2 <- tmp2[-3]
````
change the coloumn names 
````{r ,echo=TRUE}
colnames(tmp2)<-c("Date", "AvgStepsPerDay")
head(tmp2)
````
2.2 plot histogram of tmp2
````{r, histogram,echo=TRUE}
    hist(tmp2$AvgSteps, 
          main="Histogram of the total number of steps taken each day", 
          xlab = "total number of steps per day",
          border="blue", 
          col="green",
          breaks=20)
````
2.3 Calaculate the Mean and the Median respectively 
````{r ,echo=TRUE}
mean(Activity$steps,na.rm=TRUE)
median(Activity$steps,na.rm=TRUE)
````

## 3. What is the average daily activity pattern?

3.1 group the Activity dataframe by interval and store in new variable tmp3
````{r ,echo=TRUE}
tmp3<-dplyr::group_by(Activity, interval)
````

take the mean of the steps interval and store in new variable tmp4 
````{r ,echo=TRUE}
tmp4<-dplyr::summarise_each(tmp3, funs(mean(., na.rm = TRUE)))
tmp4 <- tmp4[-3]
colnames(tmp4)<-c("Interval", "AvgStepsPerInterval")
head(tmp4)
````

plot time series
````{r ,echo=TRUE}
plot(tmp4, type="l")
````

3.2 find interval which has the maximum value
````{r ,echo=TRUE}
MaxSteps<-max(tmp4$AvgSteps)
for(i in 1:288)
{
         if(tmp4$AvgSteps[i]==MaxSteps)
         {
           print(tmp4$interval[i])
         }
}
````
## 4. Imputing missing values
4.1 the number of  NAs is  values
````{r ,echo=TRUE}
sum(is.na(Activity$steps))
````

4.2 replace NAs with the average steps per interval over all days (this has already been claculated and stored in varibale tmp4)
4.3
````{r ,echo=TRUE}
Activity2<-Activity
for(i in 1: 17568)
{
  if(is.na(Activity2$steps[i]))
  {
  Activity2$steps[i]<-tmp4$AvgStepsPerInterval[which(tmp4$Interval==Activity2$interval[i])]
  }
}
head( Activity2)
tail( Activity2)
````
new mean and activity with NAs substituted with values taken for average steps per interval
````{r ,echo=TRUE}
mean(Activity2$steps)
median(Activity2$steps)
````

4.4 histogram
````{r ,echo=TRUE}
hist(Activity2$steps, 
          main="Histogram of the total number of steps taken each day",
          xlab = "total number of steps per day",
          border="blue", 
          col="green",
          breaks=30)
````

## 5. Are there differences in activity patterns between weekdays and weekends?
5.1 find the day of a week in r
````{r ,echo=TRUE}
Activity$day <- weekdays(as.Date(Activity$date))
````

group by day in variable tmp99   
````{r ,echo=TRUE}
tmp9<-dplyr::group_by(Activity, day)
tmp99<-tmp9
````

classify days as weekend or weekday
````{r ,echo=TRUE}
for(i in 1:17568)
{
  if(tmp99$day[i]=="Saturday")
  {
    tmp99$day[i]<-"Weekend"
  }
  else if(tmp99$day[i]=="Sunday")
  {
    tmp99$day[i]<-"Weekend"
  }
  else
  {
    tmp99$day[i]<-"Weekday"
  }
}
````

5.2 Take the means of each variable per activity and per day  
````{r ,echo=TRUE}
tmp10<-dplyr::summarise_each(tmp9, funs(mean(., na.rm = TRUE)))
head(tmp10)
````
